Race Science in Mainstream Psychology, 1960-2010

### 2025-04-23 paper/paper.qmd
accidentally deleted fig label
working on final formatting

bibliography: [references.yaml, refs-manual.bib]

format: 
  aft-pdf:
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    cite-method: citeproc
    include-in-header: 
        text: |
            \usepackage{lineno}
            \linenumbers
    geometry:
        - left = 1.5in
        - right = 1.5in

@@

<!-- 
QSS submission guidelines: 
<https://direct.mit.edu/qss/pages/submission-guidelines>

- 300-word cover letter on significance and fit
- Articles: original theoretical, empirical, or methodological research (typically between 5,000 and 8,000 words)

 -->

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/pf_wos_results.png){#fig-wos fig-env="figure*"}

@@

Following the approach of @HicksProductivityInterdisciplinaryImpacts2021, we did not attempt to identify a unique best fitted model for further analysis.  While the results section of this paper focuses on the medium vocabulary, $k=40$ model as the "median" among the 24 fitted models, qualitatively similar results appear across all models with $k \geq 10$ for all vocabularies; see @sec-topic-quality and figures @fig-grid-lg, @fig-grid-md, @fig-grid-sm.

To analyze the fitted topic models, we first constructed and manually inspected visualizations of the topic-document distributions (see figures @fig-gamma-lg, @fig-gamma-md, @fig-gamma-sm) and the top (highest-probability) terms for each topic (see figures @fig-silge-lg, @fig-silge-md, @fig-silge-sm).  This suggested that the models often identified two distinct topics with racial terms, that some of these topics used terms from intelligence research, and that these topics had different occurrence patterns across journals.  To examine these patterns more systematically, we programmatically identified every topic across all 24 models that used some combination of racial and intelligence terms, and assembled visualization grids of the occurrence of these topics by journal over time (see figures @fig-grid-lg, @fig-grid-md, @fig-grid-sm).

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Topics 07 and 24 are examples of the two distinct types of race science discourse, while topic 22 is "mainstream" or non-race science intelligence research.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic; text version in table @tbl-silge.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/08_focal_topics.png){#fig-focal}

Figure @fig-focal focuses on three topics identified in one of the 24 fitted topic models; see table @tbl-silge for a table version of the Silge plot in @fig-focal; figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; @fig-silge-lg; @fig-silge-md; @fig-silge-sm] for all topics from all models, and figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] for selected topics from all models; we discuss the robustness of our findings across all models below.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or (in the case of PID) these were not available for the corpus.

@@

These qualitative results are robust across models using the large and medium vocabularies; see @fig-grid-lg, @fig-grid-md, @fig-grid-sm.  Across all vocabularies, as $k$ increases, multiple different "intelligence only" topics emerge, apparently picking out different methods or aspects of intelligence research.  For example, there is often an "intelligence only" topic that is highly associated with BG.  MQ never has many publications in an "intelligence only" topic, except for $k=10, 20$ in the large vocabulary; in these two models, there is neither a "race only" nor "race-and-intelligence" topic.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually less than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). Not shown is 1963 for topic 22, where Robert A. Gordon was one of two coauthors of the single publication in that topic that year. After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Not shown here is 1963 for topic 22. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 13 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/15_presence.png){c}

@@

![Corpus assembly.](img/dataset.png){#fig-corpus width="120%" fig-env="figure*"}

![Count of documents in the corpus, by journal and year.](img/02_count.png){#fig-counts  fig-env="figure*"}

@@

We next constructed Silge plots [@SilgeTopicModeling2017], showing the top 15 (highest-probability) terms (noun phrases) from each topic in the model; see figures [@fig-silge-lg; @fig-silge-md; @fig-silge-sm].  Perusing these term lists, we focused on topics in each model that contained racial terms.  Every fitted topic model included at least one racial topic.  Importantly, in some cases these topics contained terms related to intelligence research, but in other cases they did not.

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science discourse articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models; see figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm]. In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

For the first round of review, both authors worked independently,  dichotomously coding each article as race science discourse or not.  We calculated the "false positive rate" — documents with maximum $\gamma$ in this topic that were not race science discourse — and interrater reliability using Cohen's $\kappa$.  The consensus true positive rate was 89.3% (108/121), while the consensus false positive rate was 8.3% (10/121), with 98% agreement across the two raters ($\kappa = 0.86)$.  Coding spreadsheets are included in the supplemental materials S8-S9.  For each of the 10 consensus false positive documents, their inclusion in the topic was readily explained, but in different ways for different documents.  Some included text from other documents, as when an article started in the middle of a page; others discussed perceptions of race relations or racial animus; one piece in BG discussed differences in the herding behavior of two cattle breeds, referred to as "white" and "black."

@@

@@

Figure @fig-stm shows four topics extracted from the `STM` model, automatically identified by the use of race keywords, intelligence keywords, or both, among the top 15 terms.  As in the `tmfast` analysis, the "race only" topic is almost exclusive to MQ; the "race and intelligence" topic emerges in mainstream journals before appearing in both the mainstream and MQ; and there are distinctive "intelligence only" topics.  Compare figures @fig-focal, @fig-grid-md.  One notable difference between these results is the role of PR in publishing race and intelligence.

![Silge plots and thresholded document counts for "race only," "race-and-intelligence," and "intelligence only" topics in the STM model.](img/12-stm.png){#fig-stm fig-env="figure*"}

@@

![Count of Pioneer-funded authors publishing in three focal topics, by year](img/15_pf_count.png){#fig-pf-count}

@@

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary.](img/05_lg_gamma.png){#fig-gamma-lg  fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary.](img/05_md_gamma.png){#fig-gamma-md  fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary.](img/05_sm_gamma.png){#fig-gamma-sm  fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, large vocabulary.](img/06-lg-silge.pdf){#fig-silge-lg fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](img/06-md-silge.pdf){#fig-silge-md fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](img/06-sm-silge.pdf){#fig-silge-sm fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](img/08_grid_lg.pdf){#fig-grid-lg fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](img/08_grid_md.pdf){#fig-grid-md fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](img/08_grid_sm.pdf){#fig-grid-sm fig-env="figure*"}

>>>>

bibliography: [references.yaml, refs-manual.bib]
csl: apa.csl

format: 
  aft-pdf:
    keep-tex: true
    link-citations: true
    mathspec: true
    fig.env: "figure*"
    number-sections: true
    cite-method: citeproc

@@

keywords:

- race science
- race and intelligence
- psychology
- behavior genetics
- text mining
- topic modeling

@@

![Journals publishing 6 or more Pioneer-funded researchers, Web of Science author search results.](img/fig-01.png){#fig-wos fig-env="figure*"}

@@

Following the approach of @HicksProductivityInterdisciplinaryImpacts2021, we did not attempt to identify a unique best fitted model for further analysis.  While the results section of this paper focuses on the medium vocabulary, $k=40$ model as the "median" among the 24 fitted models, qualitatively similar results appear across all models with $k \geq 10$ for all vocabularies; see @sec-topic-quality and supplements SM7, SM8, SM9<!-- figures @fig-grid-lg, @fig-grid-md, @fig-grid-sm -->.

To analyze the fitted topic models, we first constructed and manually inspected visualizations of the topic-document distributions (see figures @fig-gamma-lg, @fig-gamma-md, @fig-gamma-sm) and the top (highest-probability) terms for each topic (see supplements SM4-6<!-- figures @fig-silge-lg, @fig-silge-md, @fig-silge-sm -->).  This suggested that the models often identified two distinct topics with racial terms, that some of these topics used terms from intelligence research, and that these topics had different occurrence patterns across journals.  To examine these patterns more systematically, we programmatically identified every topic across all 24 models that used some combination of racial and intelligence terms, and assembled visualization grids of the occurrence of these topics by journal over time (see supplements SM7, SM8, SM9<!-- figures @fig-grid-lg, @fig-grid-md, @fig-grid-sm -->).

@@

![Silge plots [@SilgeTopicModeling2017] and smoothed time series for three focal topics from the medium vocabulary, $k = 40$ model.  Topics 07 and 24 are examples of the two distinct types of race science discourse, while topic 22 is "mainstream" or non-race science intelligence research.  Top row: top 15 terms by $\beta$ (term-topic distribution) for each topic; text version in table @tbl-silge.  Bottom row: count of articles associated with each topic, by journal and year.  Article counts use a threshold approach, with $\gamma$ (topic-document distribution) greater than 0.5.  Thin lines give annual values, thick lines give 5-year running averages.  Because *Behavior Genetics* and *Behavioral and Brain Sciences* are not prominent in any panel, neither is given a direct label.](img/fig-02.png){#fig-focal}

Figure @fig-focal focuses on three topics identified in one of the 24 fitted topic models; see table @tbl-silge for a table version of the Silge plot in @fig-focal; figures [@fig-gamma-lg; @fig-gamma-md; @fig-gamma-sm; and supplements SM4, SM5, SM6<!--  @fig-silge-lg; @fig-silge-md; @fig-silge-sm -->] for all topics from all models, and supplements SM7, SM8, SM9<!-- figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] --> for selected topics from all models; we discuss the robustness of our findings across all models below.  In the focal model, topic 07 is strongly associated with MQ:  MQ published dozens of articles in this topic each year, and no other journal ever published more than a handful.  The "Silge plot," showing the top 15 terms in the topic [@SilgeTopicModeling2017], contains a number of racial terms (`races`, `whites`, `negroes` and potentially `europe`, `africa`, `india`, and `japan`); as well as `book`, likely reflecting the fact that MQ published a number of book reviews, while the mainstream journals either did not publish book reviews or (in the case of PID) these were not available for the corpus.

@@

These qualitative results are robust across models using the large and medium vocabularies; see supplements SM7, SM8, SM9<!-- @fig-grid-lg, @fig-grid-md, @fig-grid-sm -->.  Across all vocabularies, as $k$ increases, multiple different "intelligence only" topics emerge, apparently picking out different methods or aspects of intelligence research.  For example, there is often an "intelligence only" topic that is highly associated with BG.  MQ never has many publications in an "intelligence only" topic, except for $k=10, 20$ in the large vocabulary; in these two models, there is neither a "race only" nor "race-and-intelligence" topic.

@@

![Disproportionate presence of Pioneer-funded authors in three focal topics.\
(A) Number of authors for three focal topics from the medium vocabulary, k= 40 model, by year. Mainstream intelligence (topic 22) shows exponential-like increase while MQ race science (topic 7) and race-and-intelligence research (topic 24) remain flat. \
(B) Pioneer-funded authors as a share (%) of all authors publishing in a given topic, by year (thin lines) and 5-year running mean (thick lines). Usually less than 15% of authors in a topic are Pioneer-funded. \
(C) Share of papers (%) with at least one Pioneer-funded author, by year (thin lines) and 5-year running mean (thick lines). Not shown is 1963 for topic 22, where Robert A. Gordon was one of two coauthors of the single publication in that topic that year. After about 1985, more than 15% of papers in topic 24 have at least one Pioneer-funded author. \
(D) Comparison of author share (panel B) to paper share (panel C). Each point is one year-topic combination. Not shown here is 1963 for topic 22. Points above the dark line $y = x$ indicate Pioneer-funded authors had a greater paper share than would be expected based on author share. Length of thin vertical lines measures "presence." \
(E) Empirical density plots of "presence," mean difference between paper share and author share for 1,000 randomly drawn sets of 13 authors. Bold vertical lines are the mean values for the Pioneer-funded authors (signed length of lines in panel D). Pioneer-funded authors are similar to randomly drawn sets for topics 7 and 22, but very different for topic 24. ](img/fig-03.png){#fig-presence}

@@

![Corpus assembly.](img/fig-S01.png){#fig-corpus width="120%" fig-env="figure*"}

![Count of documents in the corpus, by journal and year.](img/fig-S02.png){#fig-counts  fig-env="figure*"}

@@

We next constructed Silge plots [@SilgeTopicModeling2017], showing the top 15 (highest-probability) terms (noun phrases) from each topic in the model; see supplements SM4, SM5, SM6<!-- figures [@fig-silge-lg; @fig-silge-md; @fig-silge-sm] -->.  Perusing these term lists, we focused on topics in each model that contained racial terms.  Every fitted topic model included at least one racial topic.  Importantly, in some cases these topics contained terms related to intelligence research, but in other cases they did not.

The gamma distributions and Silge plots suggested that, once $k$ was sufficiently large, the topic models were distinguishing between two types of race science discourse articles.  We therefore examined the prevalence of topics that used racial terms, intelligence-research terms, or both, by journal and across all 24 models; see supplements SM7, SM8, SM9<!-- figures [@fig-grid-lg; @fig-grid-md; @fig-grid-sm] -->. In these visualizations, each row of facets corresponds to a value of $k$ (excluding 5 as too noisy), and each column of facets corresponds to one topic.  Topics are clustered based on whether the top 15 terms contain racial terms ("race"), terms related to intelligence research ("intelligence"), or both.  Line plots show the count of articles with $\gamma > 0.50$ for the given topic, by journal; thick lines are 5-year running averages and thin lines are raw counts.

@@

For the first round of review, both authors worked independently,  dichotomously coding each article as race science discourse or not.  We calculated the "false positive rate" — documents with maximum $\gamma$ in this topic that were not race science discourse — and interrater reliability using Cohen's $\kappa$.  The consensus true positive rate was 89.3% (108/121), while the consensus false positive rate was 8.3% (10/121), with 98% agreement across the two raters ($\kappa = 0.86)$.  Coding spreadsheets are included in the supplemental materials SM2 and SM3.  For each of the 10 consensus false positive documents, their inclusion in the topic was readily explained, but in different ways for different documents.  Some included text from other documents, as when an article started in the middle of a page; others discussed perceptions of race relations or racial animus; one piece in BG discussed differences in the herding behavior of two cattle breeds, referred to as "white" and "black."

@@

At the request of a reviewer, we also extracted a list of the top documents for each of the three focal topics in the $k=40$, medium vocabulary model. This list is supplement SM10.

@@

Figure @fig-stm shows four topics extracted from the `STM` model, automatically identified by the use of race keywords, intelligence keywords, or both, among the top 15 terms.  As in the `tmfast` analysis, the "race only" topic is almost exclusive to MQ; the "race and intelligence" topic emerges in mainstream journals before appearing in both the mainstream and MQ; and there are distinctive "intelligence only" topics.  Compare figure @fig-focal with supplement SM8<!-- @fig-grid-md -->.  One notable difference between these results is the role of PR in publishing race and intelligence.

![Silge plots and thresholded document counts for "race only," "race-and-intelligence," and "intelligence only" topics in the STM model.](img/fig-S07.png){#fig-stm fig-env="figure*"}

@@

![Count of Pioneer-funded authors publishing in three focal topics, by year](img/fig-S08.png){#fig-pf-count}

@@

![Fitted topic models: Gamma (topic-document) distributions, large vocabulary.](img/fig-S09.png){#fig-gamma-lg  fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, medium vocabulary.](img/fig-S10.png){#fig-gamma-md  fig-env="figure*"}

![Fitted topic models: Gamma (topic-document) distributions, small vocabulary.](img/fig-S11.png){#fig-gamma-sm  fig-env="figure*"}

<!-- 
![Silge plots (top 15 terms) for each topic and model, large vocabulary.](img/fig-S12.pdf){#fig-silge-lg fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, medium vocabulary.](img/fig-S13.pdf){#fig-silge-md fig-env="figure*"}

![Silge plots (top 15 terms) for each topic and model, small vocabulary.](img/fig-S14.pdf){#fig-silge-sm fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, large vocabulary.](img/fig-S15.pdf){#fig-grid-lg fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, medium vocabulary.](img/fig-S16.pdf){#fig-grid-md fig-env="figure*"}

![Topics with "race," "intelligence," or both occurring in the top 15 terms, small vocabulary.](img/fig-S17.pdf){#fig-grid-sm fig-env="figure*"}
 -->
